---
title: "HCA_Aging_Vignette"
author: "Abdalla Ahmed"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

In this vignette, we will determine the effect of aging on gene expression across each cell type in the intestine.

Please note that this is a DRAFT vignette - it will be updated as the integrated object is optimized. Additional functions to create figures will be added in the next iteration.

## Download the integrated h5ad object and perpare it to be ingested into seurat

```{r, message=FALSE}

####Setup wd and load google drive####
library(googledrive)
setwd("/labs/mpsnyder/abdalla/Intestine_scRNAseqAnalysis/HCA_GutWorkingGroup/h5ad_Files")
####Download the h5ad files####
#Authenticate with Google Drive
drive_auth()
#Find the folder in Google Drive
folder <- drive_get(as_id("1yLw1g-JqcLscMqCKe6_Hq1xk9woBeCkb"))
#List the contents of the folder
files <- drive_ls(folder)
#Download the folder’s contents
for (i in 2:nrow(files)) {
  drive_download(files[i, ], path = paste0("/labs/mpsnyder/abdalla/Intestine_scRNAseqAnalysis/HCA_GutWorkingGroup/h5ad_Files/", files$name[i]), overwrite = TRUE)
}

####Use Python to mainpute h5ad file for import into Seurat####
library(reticulate)

# Use the pyenv Python path
use_python("/home/ahmedabd/.pyenv/versions/3.9.7/bin/python", required = TRUE)

# Verify Python configuration
py_config()

# Import scanpy
scanpy <- import("scanpy")

# Load the .h5ad file using scanpy
adata_epi <- scanpy$read_h5ad("workshop_epi_v2_raw.h5ad")
adata_imm <- scanpy$read_h5ad("workshop_immune_v2_raw.h5ad")
adata_str <- scanpy$read_h5ad("workshop_stromal_v2_raw.h5ad")

# Extract metadata from adata
adata_metadata_epi <- py_to_r(adata_epi$obs)
adata_metadata_imm <- py_to_r(adata_imm$obs)
adata_metadata_str <- py_to_r(adata_str$obs)

# Check the structure
head(adata_metadata_epi)

# Extract data from AnnData object
counts_epi <- adata_epi$X
counts_imm <- adata_imm$X
counts_str <- adata_str$X

genes_epi <- adata_epi$var_names$to_list()
genes_imm <- adata_imm$var_names$to_list()
genes_str <- adata_str$var_names$to_list()

cells_epi <- adata_epi$obs_names$to_list()
cells_imm <- adata_imm$obs_names$to_list()
cells_str <- adata_str$obs_names$to_list()
```

## Create Seurat objects and perform pseudobulk

```{r, message=FALSE}
#### Create Seurat objects ####
library(Seurat)
seurat_obj_epi <- CreateSeuratObject(counts = t(as.matrix(counts_epi)))
seurat_obj_imm <- CreateSeuratObject(counts = t(as.matrix(counts_imm)))
seurat_obj_str <- CreateSeuratObject(counts = t(as.matrix(counts_str)))

rownames(seurat_obj_epi) <- genes_epi
rownames(seurat_obj_imm) <- genes_imm
rownames(seurat_obj_str) <- genes_str

colnames(seurat_obj_epi) <- cells_epi
colnames(seurat_obj_imm) <- cells_imm
colnames(seurat_obj_str) <- cells_str

#Add Metadata to the Seurat Object
rownames(adata_metadata_epi) <- rownames(seurat_obj_epi@meta.data)
rownames(adata_metadata_imm) <- rownames(seurat_obj_imm@meta.data)
rownames(adata_metadata_str) <- rownames(seurat_obj_str@meta.data)

seurat_obj_epi <- AddMetaData(seurat_obj_epi, metadata = adata_metadata_epi)
seurat_obj_imm <- AddMetaData(seurat_obj_imm, metadata = adata_metadata_imm)
seurat_obj_str <- AddMetaData(seurat_obj_str, metadata = adata_metadata_str)

# Verify the new metadata fields
head(seurat_obj_str@meta.data)

##Add age group metadata
library(dplyr)
metadata_epi <- seurat_obj_epi@meta.data
metadata_imm <- seurat_obj_imm@meta.data
metadata_str <- seurat_obj_str@meta.data

# Define new age categories
metadata_epi$age_group <- dplyr::case_when(
  metadata_epi$age_range == "0-1" ~ "0-1",
  metadata_epi$age_range %in% c("5-9", "5-15", "10-14", "15-19", "10-20") ~ "below_20",
  metadata_epi$age_range %in% c("20-29", "30-39") ~ "20-40",
  metadata_epi$age_range %in% c("40-49", "50-59", "60-69", "70-79", "80-89") ~ "over_40",
  TRUE ~ NA_character_  # Handle missing or undefined ranges
)

metadata_imm$age_group <- dplyr::case_when(
  metadata_imm$age_range == "0-1" ~ "0-1",
  metadata_imm$age_range %in% c("5-9", "5-15", "10-14", "15-19", "10-20") ~ "below_20",
  metadata_imm$age_range %in% c("20-29", "30-39") ~ "20-40",
  metadata_imm$age_range %in% c("40-49", "50-59", "60-69", "70-79", "80-89") ~ "over_40",
  TRUE ~ NA_character_  # Handle missing or undefined ranges
)

metadata_str$age_group <- dplyr::case_when(
  metadata_str$age_range == "0-1" ~ "0-1",
  metadata_str$age_range %in% c("5-9", "5-15", "10-14", "15-19", "10-20") ~ "below_20",
  metadata_str$age_range %in% c("20-29", "30-39") ~ "20-40",
  metadata_str$age_range %in% c("40-49", "50-59", "60-69", "70-79", "80-89") ~ "over_40",
  TRUE ~ NA_character_  # Handle missing or undefined ranges
)

# Add the new column 'age_group' back to the Seurat object
seurat_obj_epi <- AddMetaData(seurat_obj_epi, metadata = metadata_epi$age_group, col.name = "age_group")
seurat_obj_imm <- AddMetaData(seurat_obj_imm, metadata = metadata_imm$age_group, col.name = "age_group")
seurat_obj_str <- AddMetaData(seurat_obj_str, metadata = metadata_str$age_group, col.name = "age_group")

head(seurat_obj_str)
unique(seurat_obj_str@meta.data$age_range)
unique(seurat_obj_str@meta.data$author_cell_type)


####psuodobulk by author cell type across each sample ID (ignore tissue region for now)####
pseudo_bulk_epi <- AggregateExpression(seurat_obj_epi,  assays = "RNA",  return.seurat = TRUE,  # Returns a Seurat object
                                       group.by = c("author_cell_type", "sample_id", "age_group")  # Grouping by cell type and sample ID
)

pseudo_bulk_imm <- AggregateExpression(seurat_obj_imm,  assays = "RNA",  return.seurat = TRUE,  # Returns a Seurat object
                                       group.by = c("author_cell_type", "sample_id", "age_group")  # Grouping by cell type and sample ID
)

pseudo_bulk_str <- AggregateExpression(seurat_obj_str,  assays = "RNA",  return.seurat = TRUE,  # Returns a Seurat object
                                       group.by = c("author_cell_type", "sample_id", "age_group")  # Grouping by cell type and sample ID
)

tail(Cells(pseudo_bulk_str))
head(pseudo_bulk_str)

#add a column of the cell type and the age group which will serve as the categorical group variables for contrasting  (we dont add sample id as these are the biological replicates )
pseudo_bulk_epi$celltype.age <- paste(pseudo_bulk_epi$author_cell_type, pseudo_bulk_epi$age_group, sep = "_")
pseudo_bulk_imm$celltype.age <- paste(pseudo_bulk_imm$author_cell_type, pseudo_bulk_imm$age_group, sep = "_")
pseudo_bulk_str$celltype.age <- paste(pseudo_bulk_str$author_cell_type, pseudo_bulk_str$age_group, sep = "_")

tail(pseudo_bulk_str$celltype.age)
```

## DE Analysis across age

Perform DE testing on the pseudobulk data for each cell type comparing 20-39 year old donors vs 40+ year old donors

```{r, message=FALSE}
# Get unique cell types
cell_types <- unique(pseudo_bulk_str$author_cell_type)

# Initialize a list to store results
de_results <- list()

# Loop through each cell type (this can be done on epithelial and immune cell types aswell)
perform_de_analysis <- function(seurat_obj, 
                                age_group1 = "over-40", 
                                age_group2 = "20-40", 
                                test_use = "DESeq2") {
  
  # Get unique cell types
  cell_types <- unique(seurat_obj$author_cell_type)
  
  # Initialize a list to store results
  de_results <- list()
  
  # Loop through each cell type
  for (cell_type in cell_types) {
    # Create the ident.1 and ident.2 strings
    ident_1 <- paste0(cell_type, "_", age_group1)
    ident_2 <- paste0(cell_type, "_", age_group2)
    
    # Check if both groups exist in the data
    if (ident_1 %in% seurat_obj$celltype.age && ident_2 %in% seurat_obj$celltype.age) {
      # Count samples in each group
      count_1 <- sum(seurat_obj$celltype.age == ident_1)
      count_2 <- sum(seurat_obj$celltype.age == ident_2)
      
      # Print sample counts
      cat("Cell type:", cell_type, "\n")
      cat("  Samples in", ident_1, ":", count_1, "\n")
      cat("  Samples in", ident_2, ":", count_2, "\n")
      
      # Perform differential expression analysis
      de_result <- FindMarkers(object = seurat_obj, 
                               ident.1 = ident_1, 
                               ident.2 = ident_2,
                               test.use = test_use)
      
      # Store the result in the list
      de_results[[cell_type]] <- de_result
      
      # Print completion message
      cat("Completed DE analysis for", cell_type, "\n\n")
    } else {
      cat("Skipping", cell_type, "- one or both age groups not present\n\n")
    }
  }
  
  return(de_results)
}

DEresults_epi <-perform_de_analysis(pseudo_bulk_epi)
DEresults_imm <-perform_de_analysis(pseudo_bulk_imm)
DEresults_str <-perform_de_analysis(pseudo_bulk_str)

----
#(Optional)Test for one cell type at a time######

#Idents(pseudo_bulk_str) <- "celltype.age"

bulk.entero.de <- FindMarkers(object = pseudo_bulk_str, 
                            ident.1 = "Pericytes_over-40", 
                            ident.2 = "Pericytes_20-40",
                            test.use = "DESeq2")
head(bulk.entero.de, n = 15)

#Count Samples for Each Group
# Check the metadata for the groups
table(pseudo_bulk$celltype.age)
# Filter and count for Enterocytes_over-40
sum(pseudo_bulk$celltype.age == "Enterocytes_over-40")

# Filter and count for Enterocytes_20-40
sum(pseudo_bulk$celltype.age == "Enterocytes_20-40")

```
