---
title: "HGCA_SmallvsLargeIntestine_Vignette (Part 2 - in R)"
author: "Abdalla Ahmed"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

In this vignette, we will determine the differences in gene expression between the small intestine and the large intestine across each cell type in the intestine.

The input required is the pseudobulk data which is obtained from part 1 of this vignette.
## Download the integrated h5ad object and perpare it to be ingested into seurat

```{r, message=FALSE}

####load libraries####
library(DESeq2)
library(gprofiler2)
library(dplyr)
library(ggplot2)
library(purrr)
library(stringr)
library(tidyr) 
library(tibble)


####Differential analysis and pathway enrichment####
process_cell_type_region <- function(cell_type) {
  # Create safe filename (replace special characters)
  safe_cell_type <- gsub("/", "_", cell_type)
  
  # Define file paths
  counts_file <- paste0("./pseudobulk_outputs_regions/", safe_cell_type, "_pseudobulk_counts.csv")
  metadata_file <- paste0("./pseudobulk_outputs_regions/", safe_cell_type, "_metadata.csv")
  
  # Read data
  counts <- read.csv(counts_file, check.names = FALSE, , row.names = 1)
  metadata <- read.csv(metadata_file, check.names = FALSE, row.names = 1)
  
  # Skip if not enough samples
  if (nrow(metadata) < 4) {  # At least 2 donors with both regions
    message(paste("Skipping", cell_type, "- insufficient samples with both regions"))
    return(data.frame())
  }
  
  # Define the condition - now comparing regions
  metadata$Region <- factor(metadata$intestine_region)
  
  # Run DESeq2 - now comparing regions while controlling for age and other factors
  dds <- DESeqDataSetFromMatrix(
    countData = counts,
    colData = metadata,
    design = ~ Region)
  #+ sample_source + disease_ontology_term)
  
  dds <- DESeq(dds)
  res_comparison <- results(dds, contrast = c("Region", "small intestine", "large intestine"))
  res_comparison <- res_comparison[order(res_comparison$padj), ]
  
  # Perform enrichment analysis (same as before but with new direction labels)
  perform_enrichment <- function(de_genes, direction) {
    de_df <- as.data.frame(de_genes)
    
    if (direction == "up") {
      sig_genes <- rownames(de_df[de_df$padj < 0.05 & de_df$log2FoldChange > 0, ])
    } else {
      sig_genes <- rownames(de_df[de_df$padj < 0.05 & de_df$log2FoldChange < 0, ])
    }
    
    if (length(sig_genes) == 0) {
      return(data.frame())
    }
    
    enrich_go <- gost(
      query = sig_genes,
      organism = "hsapiens",
      significant = TRUE,
      sources = c("GO:BP"),
      evcodes=TRUE
    )
    
    if (is.null(enrich_go$result) || nrow(enrich_go$result) == 0) {
      return(data.frame())
    }
    
    enrich_go$result %>%
      mutate(Direction = ifelse(direction == "up", "Higher in small intestine", "Higher in large intestine"),
             CellType = cell_type)
  }
  
  up_enriched <- perform_enrichment(res_comparison, "up")
  down_enriched <- perform_enrichment(res_comparison, "down")
  
  up_enriched <- up_enriched[up_enriched$term_size >= 5 & up_enriched$term_size <= 350, ]
  down_enriched <- down_enriched[down_enriched$term_size >= 5 & down_enriched$term_size <= 350, ]
  
  return(bind_rows(up_enriched, down_enriched))
}

# Get all cell type files by region
cell_type_files <- list.files(path = "./pseudobulk_outputs_regions/", 
                              pattern = "_pseudobulk_counts.csv",
                              full.names = FALSE)

# Process all cell types and combine results (by region)
all_enrichments <- map_dfr(cell_types, ~ {
  result <- tryCatch(
    process_cell_type_region(.x),
    error = function(e) {
      message("Error processing ", .x, ": ", e$message)
      return(data.frame())
    }
  )
  if (nrow(result) > 0) result else NULL
})

saveRDS(all_enrichments, "all_enrichments_region(nobatchcorrection).rds")

```

## Assign pathways to categories and create figure

```{r, message=FALSE}
all_enrichments <- readRDS("all_enrichments_region(nobatchcorrection).rds")
all_enrichments <- all_enrichments |> dplyr::select(-parents)

#Note: the assignments below were created through manual curation and review of all the differential pathways across each cell type
category_keywords <- list(
  Immune_Response = c("immune", "inflamm", "interferon", "cytokine", "chemokine", "leukocyte",  "phagocytosis", "interleukin", "NF-kappaB", "lipopolysaccharide", "bacteria", "antigen", "viral",
                      "lymphocyte", "T cell", "B cell", "myeloid", "antigen", "defense", "complement", "MHC"),
  Metabolism = c("insulin", "metaboli", "biosynthe", "catabolic", "lipid", "fatty acid", "steroid", "glucose",
                 "cholesterol", "amino", "organic acid", "carboxylic acid", "triglyceride", "saccharide"),
  Mitochondrial_Energy = c("oxidative phosphorylation", "mitochondr", "respiration", 
                           "electron transport", "ATP synthesis", "aerobic respiration", "electron", "proton", "energy", "respiratory chain complex"),
  Digestion_and_Absorption = c("digest", "absorption", "intestinal"),
  Transport_And_Secretion = c("transport", "secretion", "vesicle", "exocytosis", "endocytosis", 
                              "golgi", "endosome", "ion transport", "trafic"),
  Cell_Cycle_And_Division = c("cell cycle", "mitotic", "mitosis", "replication", "division", "centrosome",
                              "spindle", "segregation", "anaphase", "telophase", "duplication", "separation"),
  Stress_Response = c("stress", "apopto", "repair", "hypoxia", "oxygen", "shear", "protein folding", "chaperone", "unfolded protein", "telomere", "ubiquitination", "stabili", "refolding", "oxidative", "reactive oxygen"),
  Cell_Adhesion_And_Matrix = c("cell adhesion", "extracellular matrix", "collagen", 
                               "integrin", "junction"),
  Protein_Glycosylation = c("glycan", "glycosylation"),
  Translation = c("ribosom", "translat", "rRNA")
)


# Function to assign categories
assign_category <- function(term, categories) {
  found <- sapply(names(categories), function(cat) {
    any(sapply(categories[[cat]], function(keyword) grepl(keyword, term, ignore.case = TRUE)))
  })
  if (any(found)) {
    # Return first matching category, or paste all matches if multiple wanted
    names(categories)[which(found)[1]]
  } else {
    "Other"
  }
}

unique_terms <- unique(all_enrichments$term_name)

category <- sapply(unique_terms, assign_category, categories = category_keywords)
write.csv(category, "top_pathways_cat.csv", row.names = TRUE)

category <- read.csv("top_pathways_cat.csv")
Master_Cell_Annotation <- read.csv("unique_prelim_annotations.csv")

# Add 'master' cell type annotation to all_enrichments
all_enrichments <- all_enrichments %>%
  filter(CellType != "doublet") %>%
  #  filter(CellType != "Epithelial") %>%
  filter(CellType != "lowQ") %>%
  # Replace underscores with spaces for matching
  mutate(CellType_clean = str_replace_all(CellType, "_", " ")) %>%
  left_join(Master_Cell_Annotation %>% 
              dplyr::select(hgca_celltype_v1, cell_type), 
            by = c("CellType_clean" = "hgca_celltype_v1")) %>%
  dplyr::select(-CellType_clean) 

all_enrichments_toSave <- all_enrichments %>%
  dplyr::select(-query, -evidence_codes) 
write.csv(all_enrichments_toSave, "all_enrichments_HGCApaper.csv", row.names = TRUE)


top_n_per_celltype <- 1  # Number of top pathways to show per cell type per direction

top_pathways <- all_enrichments %>% group_by(CellType, Direction) %>% arrange(p_value) %>%
  slice_head(n = top_n_per_celltype) %>% ungroup() %>% distinct(term_name) %>% 
  as.data.frame()

all_terms_with_categories <- all_enrichments %>% distinct(term_name) %>%
  mutate(category = sapply(term_name, assign_category, categories = category_keywords))

# Count how many pathways we have per category in our initial selection
category_counts <- top_pathways %>%
  left_join(all_terms_with_categories, by = "term_name") %>% group_by(category) %>%
  summarise(n = n()) %>% ungroup()

all_categories <- names(category_keywords)

categories_needing_more <- all_categories[!all_categories %in% category_counts$category | 
                                            all_categories %in% category_counts$category[category_counts$n < 3]]

term_frequencies <- all_enrichments %>% 
  dplyr::count(term_name, name = "n")

supplemental_pathways <- map_dfr(categories_needing_more, function(cat) {
  # How many are we missing? (3 if new category, 3-current if existing)
  needed <- ifelse(cat %in% category_counts$category, 
                   3 - category_counts$n[category_counts$category == cat],
                   3)
  
  all_terms_with_categories %>%
    filter(category == cat) %>%
    filter(!term_name %in% top_pathways$term_name) %>%
    left_join(term_frequencies, by = "term_name") %>%
    arrange(desc(n)) %>%
    slice_head(n = needed)
})

top_pathways <- unique(c(top_pathways$term_name, supplemental_pathways$term_name))

#create the figure
library(ggplot2)

# 1. Project palette (Wong colorblind-safe + HCA blue)
PAL_GCA <- c(
  hgca       = "#0072B2",  # HCA blue (Wong "Blue")
  taurus     = "#D55E00",  # Vermillion
  shared     = "#009E73",  # Bluish green
  null       = "#E0E0E0",  # Light grey
  text       = "#000000"
)

# 2. Theme: Helvetica, no gridlines, thin axes, sentence-case labels.
theme_gca <- function(base_size = 6, base_family = "Helvetica") {
  theme_classic(base_size = base_size, base_family = base_family) +
    theme(
      text             = element_text(family = base_family, color = "black"),
      plot.title       = element_text(family = base_family, face = "bold",
                                      size = 7, color = "black", hjust = 0),
      axis.line        = element_line(color = "black", linewidth = 0.25),
      axis.ticks       = element_line(color = "black", linewidth = 0.25),
      axis.text        = element_text(color = "black", size = 6),
      axis.title       = element_text(color = "black", size = 6),
      panel.grid       = element_blank(),
      panel.background = element_blank(),
      plot.background  = element_blank(),
      legend.position  = "right",
      legend.key       = element_blank(),
      legend.text      = element_text(color = "black", size = 6),
      legend.title     = element_text(color = "black", size = 6, face = "bold"),
      strip.background = element_blank(),
      strip.text       = element_text(color = "black", size = 6, face = "bold")
    )
}

# 3. Save as Illustrator-editable PDF + SVG + 300 dpi PNG preview.
save_figure <- function(p, base, width_mm, height_mm) {
  width_in  <- width_mm  / 25.4
  height_in <- height_mm / 25.4
  ggplot2::ggsave(paste0(base, ".pdf"), p,
                  width = width_in, height = height_in,
                  device = grDevices::cairo_pdf)
  ggplot2::ggsave(paste0(base, ".svg"), p,
                  width = width_in, height = height_in,
                  device = svglite::svglite)
  ggplot2::ggsave(paste0(base, ".png"), p,
                  width = width_in, height = height_in, dpi = 300)
}


{
  terms_to_keep <- category %>% filter(x != "Other") %>% pull(X)
  
  enrichments_filtered <- all_enrichments %>%
    filter(term_name %in% terms_to_keep) %>%
    filter(term_name %in% top_pathways)
  
  if (nrow(enrichments_filtered) == 0) {
    message("No enrichments found")
  } else {
    
    # ---- master cell type ordering ----
    desired_cell_type_order <- c("epithelial", "stroma", "myeloid", "lymphoid")
    
    master_celltype_colors <- c(
      "epithelial" = "#A65628", "stroma" = "#377EB8",
      "myeloid" = "#4DAF4A", "lymphoid" = "#FF7F00"
    )
    
    # ---- build plot_data ----
    plot_data <- enrichments_filtered %>%
      mutate(
        Direction_Label = ifelse(Direction == "Higher in small intestine",
                                 "Higher in small intestine", "Higher in large intestine"),
        Enrichment = -log10(p_value),
        shape = ifelse(Direction == "Higher in small intestine", 24, 25),
        cell_type = factor(cell_type, levels = desired_cell_type_order)
      )
    
    # ---- build CellType order (your original logic) ----
    cell_types_ordered <- plot_data %>%
      arrange(cell_type, CellType) %>%
      pull(CellType) %>%
      unique()
    
    celltype_mapping <- data.frame(
      CellType = cell_types_ordered,
      cell_type_group = sapply(cell_types_ordered, function(ct) {
        plot_data %>%
          filter(CellType == ct) %>%
          pull(cell_type) %>%
          as.character() %>%
          unique() %>%
          dplyr::first()
      })
    ) %>%
      mutate(x_position = 1:n())
    
    vline_positions <- celltype_mapping %>%
      group_by(cell_type_group) %>%
      summarize(last_x = max(x_position), .groups = "drop") %>%
      pull(last_x) + 0.5
    
    vline_positions <- sort(vline_positions)
    vline_positions <- vline_positions[-length(vline_positions)]
    
    # ---- add category + set category order ----
    category_order <- c(
      "Digestion_and_Absorption", "Transport_And_Secretion", "Metabolism", "Mitochondrial_Energy",
      "Immune_Response", "Stress_Response", "Cell_Cycle_And_Division", "Cell_Adhesion_And_Matrix",
      "Protein_Glycosylation", "Translation"
    )
    
    plot_data <- plot_data %>%
      left_join(category %>% select(X, x), by = c("term_name" = "X")) %>%
      mutate(category = factor(x, levels = category_order))
    
    # ---- (A) keep your column clustering within each master cell type ----
    unique_celltypes <- plot_data %>%
      distinct(CellType, cell_type) %>%
      arrange(cell_type, CellType)
    
    celltype_order_clustered <- unique_celltypes %>%
      group_by(cell_type) %>%
      nest() %>%
      mutate(
        clustered_order = map(data, ~ {
          mat <- plot_data %>%
            filter(CellType %in% .x$CellType) %>%
            group_by(term_name, CellType) %>%
            summarize(avg_enrich = mean(Enrichment, na.rm = TRUE), .groups = "drop") %>%
            pivot_wider(names_from = CellType, values_from = avg_enrich, values_fill = 0) %>%
            column_to_rownames("term_name") %>%
            as.matrix()
          
          if (ncol(mat) > 1) hclust(dist(t(mat)), method = "complete")$order else 1
        }),
        ordered_celltypes = map2(data, clustered_order, ~ .x$CellType[.y])
      ) %>%
      unnest(ordered_celltypes) %>%
      pull(ordered_celltypes) %>%
      unique()
    
    plot_data <- plot_data %>%
      mutate(CellType = factor(CellType, levels = celltype_order_clustered))
    
    # ---- (B) cluster ROWS (term_name) within each category ----
    term_order_clustered <- plot_data %>%
      filter(!is.na(category)) %>%
      group_by(category) %>%
      nest() %>%
      mutate(
        term_levels = map(data, ~ {
          df_cat <- .x
          
          # term x CellType matrix (mean enrichment); fill missing as 0
          mat <- df_cat %>%
            group_by(term_name, CellType) %>%
            summarize(avg_enrich = mean(Enrichment, na.rm = TRUE), .groups = "drop") %>%
            pivot_wider(names_from = CellType, values_from = avg_enrich, values_fill = 0) %>%
            column_to_rownames("term_name") %>%
            as.matrix()
          
          if (nrow(mat) <= 1) return(rownames(mat))
          
          ord <- hclust(dist(mat), method = "complete")$order
          rownames(mat)[ord]
        })
      ) %>%
      arrange(category) %>%
      pull(term_levels) %>%
      unlist(use.names = FALSE) %>%
      unique()
    
    plot_data <- plot_data %>%
      mutate(term_name = factor(term_name, levels = term_order_clustered))
    
    # ---- plot ----
    heatmap_plot <- ggplot(plot_data, aes(x = CellType, y = term_name)) +
      geom_point(aes(fill = Direction_Label, size = Enrichment, shape = shape),
                 stroke = 0.25, na.rm = TRUE) +
      geom_vline(xintercept = vline_positions,
                 color = "black", linewidth = 0.25, linetype = "solid") +
      scale_shape_identity() +
      scale_fill_manual(
        values = c(
          "Higher in small intestine" = "#0072B2",  # Vermillion from PAL_GCA
          "Higher in large intestine" = "#D55E00"   # HCA blue from PAL_GCA
        )
      ) +
      scale_size_continuous(range = c(0.5, 3), name = "-log10(p-value)") +
      facet_grid(category ~ ., scales = "free_y", space = "free_y") +
      theme_gca() +
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title = element_blank(),
        panel.spacing.y = unit(0.3, "lines"),
        #legend.position = "none"
      ) +
      guides(
        fill = guide_legend(override.aes = list(shape = 24, size = 2), order = 1),
        size = guide_legend(override.aes = list(shape = 24, fill = "grey50"), order = 2)
      )
    
    save_figure(
      heatmap_plot,
      base = "Figures/ALLcategories_Pathway_Enrichment_Heatmap_clusterRowsWithinCategory(hgca)",
      width_mm = (1.6*180),
      height_mm = (1.6*170)
    )
    
    message("Plot saved successfully")
  }
}


```

